Machine Learning Models for the Segmentation of E-coaching Text

### 2018-03-07 paper-amia-2018/manuscript_before_1st_feedback.tex
update introduction after April's feedback

\author{Mehedi Hasan, BS$^{1}$\footnote[1]{Authors provided equal contribution. \label{footnote1}}, Alexander Kotov, PhD$^{1}$\textsuperscript{\ref{footnote1}}, April Idalski Carcone, PhD$^{2}$, Ming Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$}

\institutes{
$^1$Department of Computer Science, Wayne State University, Detroit, Michigan \\  
$^2$Department of Family Medicine and Public Health Sciences, School of Medicine, Wayne State University, Detroit, Michigan\\
}

@@

\section*{Introduction}
Unhealthy eating habits, particularly low fruit and vegetable intake, is a growing, serious public health concern, particularly among young adults age 21-30, referred to as Generation Y (GenY)\cite{blanck2008trends,centers2007fruit}. This generation has adopted a lifestyle that involves eating accessible, ``no mess'', quick, ``grab and go'' foods\cite{nebeling2007still,brug1999application}. They mainly eat ``out'' and infrequently shop and prepare food, limiting access to fruit and vegetables (FV)\cite{nelson2009improving,larson2006food}. Unfortunately, less than one-third of US adults\cite{blanck2008trends,ogden2006prevalence} and only 20\% of GenY\cite{blanck2008trends,american2006american,thompson2005dietary} eat the recommended 5 servings of fruit and vegetables daily. Those in inner-city urban and rural settings have among the poorest eating habits\cite{blanck2008trends,centers2007fruit,ogden2006prevalence,american2006american,thompson2005dietary}. GenY's poor dietary practices placing them at high risk for obesity and many chronic diseases, such as type 2 diabetes, as well as declines in predicted health status and life expectancy. Thus, there is a need to develop effective interventions to improve GenY's eating habits.

@@

A major drawback of this research is the qualitative methods traditionally used to analyze the communication process which is resource-intensive, requiring an iterative process of human (subjective) interpretation of the text. Rapidly developing computational technologies, specifically, machine learning models, offer a unique opportunity to accelerate this process. Our research group has recently applied machine learning-based models to MI sessions \cite{hasan2016study,kotov2015interpretable}. A simple communication code scheme was automated to characterize patient communication and achieved accuracy comparable to human coders\cite{hasan2016study}. However, this study utilized e-coaching data which have different format and context compared to motivational interview transcripts. MI session data was prepared by transcribing audio conversation which involves a counselor, an adolescent, and a caregiver. On the other hand, e-coaching data was directly collected from the email responses between an eCoach and young adult. By utilizing this e-coaching data and leveraging innovative machine learning models, the ultimate goal of this research study is to fully automate the communication coding process in eCoach-patient communication to increases in fruit and vegetable intake.

However, a significant barrier to fully automate e-coaching is the unsegmented email exchange. E-coaching data comprised of email responses which need to be segmented into groups of MI behaviors for developing an automatic classification of clinical interactions. However, automatic segmentation of e-coaching intervention sessions is a challenging task due to the 2 important reasons. First, the email is an unstructured text that contains informal email exchange in non-traditional formats. Second, a text segment not necessarily belongs to the entire sentence or collection of sentences. One sentence can be segmented into several MI behaviors and vice versa. Figure~\ref{fig:text-segment} illustrates the segmentation of an e-coaching email exchange where the first sentence segmented into 2 different MI behaviors. On the other hand, fourth and fifth segments contain only one and multiple sentences, respectively.

@@

We observed worst results of RNN, in particular, LSTM and GRU for the text segmentation. We believe that RNN performed poorly because it has a large set of weights which required a large set of data for both classes. In this study, we utilized 3,138 examples of boundary case which failed to achieve good results. While the performance of RNN is poor, GRU performs better than LSTM which was observed in other previous study\cite{chung2014empirical}.

>>>>

\author{Mehedi Hasan, BS$^{1}$\footnote[1]{Authors provided equal contribution. \label{footnote1}}, Alexander Kotov, PhD$^{1}$\textsuperscript{\ref{footnote1}}, Ming Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$, Gwen L. Alexander, PhD$^{3}$, April Idalski Carcone, PhD$^{4}$}

\institutes{
$^1$Department of Computer Science, Wayne State University, Detroit, Michigan \\  
$^2$Director, Center for Translational Behavioral Research, Department of Behavioral Sciences and Social Medicine, Florida State University, Tallahassee, Florida\\
$^3$Department of Public Health Sciences, Henry Ford Health System, Detroit, Michigan\\
$^4$Department of Family Medicine and Public Health Sciences, School of Medicine, Wayne State University, Detroit, Michigan\\
}

@@

\section*{Introduction}
Unhealthy eating habits, particularly low fruit and vegetable intake, is a growing, serious public health concern, particularly among young adults age 21-30, referred to as Generation Y (GenY)\cite{blanck2008trends,centers2007fruit}. GenY has adopted a lifestyle that involves eating accessible, ``no mess'', quick, ``grab and go'' foods\cite{nebeling2007still,brug1999application}. They mainly eat ``out'' and infrequently shop and prepare food, limiting access to fruit and vegetables (FV)\cite{nelson2009improving,larson2006food}. Unfortunately, less than one-third of US adults\cite{blanck2008trends,ogden2006prevalence} and only 20\% of GenY\cite{blanck2008trends,american2006american,thompson2005dietary} eat the recommended 5 servings of fruit and vegetables daily. Those in inner-city urban and rural settings have among the poorest eating habits\cite{blanck2008trends,centers2007fruit,ogden2006prevalence,american2006american,thompson2005dietary}. GenY's poor dietary practices place them at high risk for obesity and many chronic diseases, such as type 2 diabetes, as well as declines in predicted health status and life expectancy. Thus, there is a need to develop effective interventions to improve GenY's eating habits.

@@

A major drawback will be the qualitative methods traditionally used to analyze the communication process. Qualitative coding of communication data is resource-intensive, requiring an iterative process of human (subjective) interpretation of the text. Rapidly developing computational technologies, specifically, machine learning models, offer a unique opportunity to accelerate this process. Our research group has recently applied machine learning-based models to MI sessions \cite{hasan2016study,kotov2015interpretable}. A simple communication code scheme was automated to characterize patient communication and achieved accuracy comparable to human coders\cite{hasan2016study}. However, this study utilized e-coaching data which have different format and context compared to motivational interview transcripts. MI session data was prepared by transcribing audio conversation which involves a counselor, an adolescent, and a caregiver. On the other hand, e-coaching data was directly collected from the email exchange between an eCoach and young adult. By utilizing this e-coaching data and leveraging innovative machine learning models, the ultimate goal of this research study is to fully automate the communication coding process in eCoach-patient communication to increases in fruit and vegetable intake.

A significant barrier to fully automated text classification is the segmentation of the text data. During traditional qualitative coding, coders determine where to segment text, that is, where one code ends and another begins. E-coaching data is comprised of email responses which are unsegmented, unlike more traditional diadic clinical communication where segmentation occurs naturally due to it's conversation nature. Automatic segmentation of e-coaching intervention sessions is a challenging task due to the 2 important reasons. First, the email is an unstructured text that contains informal email exchange in non-traditional formats. Second, a text segment does not necessarily belong to the entire sentence or collection of sentences. One sentence can be segmented into multiple MI behaviors or several sentences may represent a single MI code. Figure~\ref{fig:text-segment} illustrates the segmentation of an e-coaching email exchange where the first sentence segmented into 2 different MI behaviors. On the other hand, fourth and fifth segments contain only one and multiple sentences, respectively.

@@

We observed worst results of RNN, in particular, LSTM and GRU for the text segmentation. We believe that RNN performed poorly because it has a large set of weights which required longer sequence with a large set of data for both classes. In this study, maximum length of the input sequence was 3 and we utilized 3,138 examples of boundary case which failed to achieve good results. While the performance of RNN is poor, GRU performs better than LSTM which was observed in other previous study\cite{chung2014empirical}.
